---
title: "LadybugDB extensions from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LadybugDB extensions from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

LadybugDB extensions add graph algorithms, indexes, and external data sources
without increasing the core engine size. Extension installation downloads
platform-specific code, so the examples are not executed while this vignette
is built.

```{r eval=FALSE}
library(rladybugdb)

db <- lb_database("extensions.lbug")
conn <- lb_connection(db)
on.exit({
  lb_close(conn)
  lb_close(db)
}, add = TRUE)

lb_list_extensions(conn)
lb_close(lb_install_extension(conn, "algo"))
lb_close(lb_load_extension(conn, "algo"))
```

The helpers validate official extension names and issue the same `INSTALL`,
`LOAD`, and `UPDATE` commands available directly through `lb_execute()`.

## PageRank and Louvain

Algorithms run on a projected graph. Given `Person` nodes and `KNOWS`
relationships:

```{r eval=FALSE}
lb_close(lb_execute(
  conn,
  "CALL project_graph('SocialGraph', ['Person'], ['KNOWS'])"
))

page_rank <- lb_query(
  conn,
  paste(
    "CALL page_rank('SocialGraph')",
    "RETURN node.name AS name, rank AS pagerank",
    "ORDER BY rank DESC"
  )
)

communities <- lb_query(
  conn,
  paste(
    "CALL louvain('SocialGraph')",
    "RETURN louvain_id, collect(node.name) AS members",
    "ORDER BY louvain_id"
  )
)
```

## BM25 full-text search

The `FTS` extension indexes string properties on node tables and ranks matches
with Okapi BM25.

```{r eval=FALSE}
lb_close(lb_install_extension(conn, "fts"))
lb_close(lb_load_extension(conn, "fts"))
lb_close(lb_execute(
  conn,
  paste0(
    "CALL CREATE_FTS_INDEX(",
    "'Book', 'book_index', ['title', 'abstract'], stemmer := 'porter')"
  )
))

hits <- lb_query(
  conn,
  paste(
    "CALL QUERY_FTS_INDEX('Book', 'book_index', $query, top := 10)",
    "RETURN node.title AS title, score ORDER BY score DESC"
  ),
  parameters = list(query = "quantum machine")
)
```

## HNSW vector similarity search

The `vector` extension provides a native disk-backed HNSW index over fixed
FLOAT or DOUBLE arrays.

```{r eval=FALSE}
lb_close(lb_install_extension(conn, "vector"))
lb_close(lb_load_extension(conn, "vector"))
lb_close(lb_execute(
  conn,
  paste0(
    "CALL CREATE_VECTOR_INDEX(",
    "'Book', 'book_title_index', 'title_embedding', metric := 'cosine')"
  )
))

nearest <- lb_query(
  conn,
  paste(
    "CALL QUERY_VECTOR_INDEX('Book', 'book_title_index', $embedding, 5)",
    "RETURN node.title AS title, distance ORDER BY distance"
  ),
  parameters = list(embedding = as.numeric(query_embedding))
)
```

## DuckDB, PostgreSQL, and SQLite

Install and load the corresponding extension, attach the external database,
then scan its tables with `LOAD FROM`. Credentials and paths below are
placeholders and must come from trusted configuration.

```{r eval=FALSE}
for (extension in c("duckdb", "postgres", "sqlite")) {
  lb_close(lb_install_extension(conn, extension))
  lb_close(lb_load_extension(conn, extension))
}

lb_close(lb_execute(conn, "ATTACH 'warehouse.duckdb' AS wh (dbtype duckdb)"))
duck_rows <- lb_query(conn, "LOAD FROM wh.events RETURN * LIMIT 100")

lb_close(lb_execute(
  conn,
  "ATTACH 'host=localhost dbname=app' AS pg (dbtype postgres)"
))
postgres_rows <- lb_query(conn, "LOAD FROM pg.public.events RETURN * LIMIT 100")

lb_close(lb_execute(conn, "ATTACH 'app.sqlite' AS sq (dbtype sqlite)"))
sqlite_rows <- lb_query(conn, "LOAD FROM sq.events RETURN * LIMIT 100")
```

## Parquet, Iceberg, and ADBC

Parquet can be scanned directly by the core import syntax. Iceberg and ADBC
use extensions and expose their sources through the same `LOAD FROM` model.

```{r eval=FALSE}
parquet_rows <- lb_query(conn, "LOAD FROM 'events/*.parquet' RETURN *")

lb_close(lb_install_extension(conn, "iceberg"))
lb_close(lb_load_extension(conn, "iceberg"))
# Configure the catalog according to the LadybugDB Iceberg documentation,
# then scan the attached catalog.table with LOAD FROM.

lb_close(lb_install_extension(conn, "adbc"))
lb_close(lb_load_extension(conn, "adbc"))
# Install the database vendor's ADBC driver, attach it using that driver's
# connection options, then query its tables with LOAD FROM.
```

Extension availability and parameters follow the installed LadybugDB engine.
Use `lb_list_extensions()` and the upstream extension documentation as the
source of truth for a deployment.
