---
title: "Getting Started with rladybugdb"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rladybugdb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(rladybugdb)
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = ladybugdb_is_installed()
)
```

## Introduction

`rladybugdb` provides an R interface to **LadybugDB**, an embedded columnar
graph database that uses the **openCypher** query language. LadybugDB is a fork
of [KuzuDB](https://kuzudb.com) and supports the same Cypher dialect.

Starting with v0.2.0 the package uses **native Rcpp C++ bindings** compiled
against the LadybugDB C library (`liblbug`). The C library is bundled with the
package — no Python, no `reticulate`, and no extra installation step are
required.

```
R → Rcpp (rladybugdb.so) → C API (lbug.h) → LadybugDB engine
```

## Installation

Install from GitHub:

```r
# install.packages("remotes")
remotes::install_github("hadimaster65555/rladybugdb")
```

The `configure` script downloads the prebuilt `liblbug` binary for your
platform at install time. After that the package is fully self-contained.

Verify the library is working:

```{r version}
library(rladybugdb)
ladybugdb_version()      # e.g. "0.15.2"
ladybugdb_is_installed() # TRUE
```

> **Note:** `ladybugdb_install()` is deprecated and no longer does anything.
> You do not need to call it.

## Opening a Database

`lb_database()` opens or creates a LadybugDB database. Pass `":memory:"` for a
transient in-memory database (ideal for testing and interactive exploration):

```{r open-db}
library(rladybugdb)

db   <- lb_database(":memory:")
conn <- lb_connection(db)
db
conn
```

For a persistent on-disk database pass a directory path:

```r
db <- lb_database("~/my_graph_db")
```

Use `lb_close()` to release resources when done:

```r
lb_close(conn)
lb_close(db)
```

## Defining a Schema

LadybugDB uses a **property graph** model: nodes belong to *node tables* and
edges belong to *relationship tables*. Both are defined with Cypher DDL:

```{r schema}
# Node tables
lb_execute(conn, "CREATE NODE TABLE Person (
  name   STRING,
  age    INT64,
  PRIMARY KEY (name)
)")

lb_execute(conn, "CREATE NODE TABLE City (
  name    STRING,
  country STRING,
  PRIMARY KEY (name)
)")

# Relationship table (directed: Person → City)
lb_execute(conn, "CREATE REL TABLE LivesIn (FROM Person TO City, since INT64)")
```

## Inserting Data

### Row-by-row with Cypher `CREATE`

```{r insert}
lb_execute(conn, "CREATE (:Person {name: 'Alice', age: 30})")
lb_execute(conn, "CREATE (:Person {name: 'Bob',   age: 25})")
lb_execute(conn, "CREATE (:City   {name: 'London', country: 'UK'})")
lb_execute(conn, "CREATE (:City   {name: 'Paris',  country: 'France'})")

lb_execute(conn, "
  MATCH (p:Person {name: 'Alice'}), (c:City {name: 'London'})
  CREATE (p)-[:LivesIn {since: 2018}]->(c)
")
lb_execute(conn, "
  MATCH (p:Person {name: 'Bob'}), (c:City {name: 'Paris'})
  CREATE (p)-[:LivesIn {since: 2021}]->(c)
")
```

### Bulk loading from an R data frame

For larger datasets `lb_copy_from_df()` writes the data frame to a temporary
CSV and uses LadybugDB's `COPY … FROM` loader — much faster than individual
`CREATE` statements:

```{r bulk-load}
more_people <- data.frame(
  name = c("Carol", "Dave"),
  age  = c(35L,     28L)
)
lb_copy_from_df(conn, more_people, "Person")

lb_query(conn, "MATCH (p:Person) RETURN p.name AS name, p.age AS age ORDER BY p.name")
```

### Loading from a CSV file

```r
lb_copy_from_csv(conn, "/data/people.csv", "Person")
```

## Querying Data

### `lb_query()` — query directly to a data frame

```{r query-df}
df <- lb_query(conn,
  "MATCH (p:Person)-[:LivesIn]->(c:City)
   RETURN p.name AS person, c.name AS city, c.country AS country
   ORDER BY p.name")
df
```

### `lb_execute()` — query to an `lb_result`

`lb_result` holds a lazy cursor. Convert it explicitly:

```{r query-execute}
result <- lb_execute(conn,
  "MATCH (p:Person) RETURN p.name AS name, p.age AS age ORDER BY p.age")
result          # prints column schema
as.data.frame(result)
```

### Parameterised queries

Use named parameters (prefixed with `$`) instead of string interpolation.
This avoids Cypher-injection bugs and handles type coercion automatically:

```{r params}
lb_execute(conn,
  "MATCH (p:Person {name: $name}) RETURN p.age AS age",
  parameters = list(name = "Alice"))
```

Supported parameter types: `character`, `integer`, `double`, `logical`.

### As a tibble

```{r query-tibble, eval = ladybugdb_is_installed() && requireNamespace("tibble", quietly = TRUE)}
result <- lb_execute(conn,
  "MATCH (p:Person) RETURN p.name AS name, p.age AS age ORDER BY p.age")
tibble::as_tibble(result)
```

### As an Arrow Table

```{r query-arrow, eval = ladybugdb_is_installed() && requireNamespace("arrow", quietly = TRUE)}
result <- lb_execute(conn,
  "MATCH (p:Person) RETURN p.name AS name, p.age AS age")
as_arrow_table(result)
```

## Graph Analysis

When a query returns full node and relationship columns (i.e. `RETURN p, r, c`
rather than individual properties), `as_igraph()` and `as_tbl_graph()` convert
the result into graph objects.

```{r igraph, eval = ladybugdb_is_installed() && requireNamespace("igraph", quietly = TRUE)}
result <- lb_execute(conn,
  "MATCH (p:Person)-[r:LivesIn]->(c:City) RETURN p, r, c")
g <- as_igraph(result)
print(g)
```

```{r tidygraph, eval = ladybugdb_is_installed() && requireNamespace("tidygraph", quietly = TRUE)}
result2 <- lb_execute(conn,
  "MATCH (p:Person)-[r:LivesIn]->(c:City) RETURN p, r, c")
tg <- as_tbl_graph(result2)
tg
```

## Cleanup

```{r cleanup}
lb_close(conn)
lb_close(db)
```

## Type Mapping Reference

| LadybugDB type | R type | Notes |
|---|---|---|
| `INT8` / `INT16` / `INT32` | `integer` | |
| `INT64` / `SERIAL` | `double` | Exact up to 2^53; no truncation for graph IDs |
| `FLOAT` / `DOUBLE` | `double` | |
| `BOOLEAN` | `logical` | |
| `STRING` / `UUID` | `character` | |
| `DATE` | `Date` | Days since Unix epoch |
| `TIMESTAMP` | `POSIXct` | Microseconds ÷ 1e6 → seconds since epoch |
| `INTERVAL` / `DECIMAL` / `BLOB` | `character` | Serialised as string |
| `NULL` | `NA` | Typed NA matching the column type |
| `LIST` / `ARRAY` | list column | Each cell is an R `list` |
| `MAP` | `list` with `$keys` / `$values` | |
| `STRUCT` | named `list` | |
| `NODE` | named `list` (`_ID`, `_LABEL`, properties) | Use `as_igraph()` to unwrap |
| `REL` | named `list` (`_SRC`, `_DST`, `_LABEL`, `_ID`, properties) | |

## Real-world example: OpenFlights

`example_openflights.R` (in the package root) shows a complete workflow with
real data:

1. **Download** OpenFlights airport and route CSVs (~6 000 airports, ~66 000 routes)
2. **Load** into LadybugDB with `lb_copy_from_df()`
3. **Query** top hubs, country rankings, and a hub subgraph using Cypher
4. **Visualise** with `ggplot2`, `ggraph`, and the `maps` package

```r
# From the package source directory:
Rscript example_openflights.R
```

Plots are written to `example_openflights_plots/`.
