---
title: "Stored databases and engine migration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stored databases and engine migration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`rladybugdb` 0.4.0 embeds LadybugDB 0.20.4. Earlier native releases of the
package embedded LadybugDB 0.15.2. LadybugDB 0.20.4 can read the storage
version produced by 0.15.x; the package tests this with a database created by
the old engine rather than only testing a newly created file.

## Safe upgrade workflow

Treat a database file as application data, not as a disposable package cache.
Before opening it with a newer engine:

1. Stop all processes that can write to the database.
2. Copy the database file and any associated WAL file to a backup location.
3. Open the copied database with the new package and run representative
   read-only validation queries.
4. Open the production copy for writing only after those checks pass.

LadybugDB 0.20 uses a single database file, so pass a file path rather than a
directory path:

```{r eval=FALSE}
library(rladybugdb)

db <- lb_database("analytics.lbug", read_only = TRUE)
conn <- lb_connection(db)
on.exit({
  lb_close(conn)
  lb_close(db)
}, add = TRUE)

lb_query(conn, "CALL SHOW_TABLES() RETURN *")
lb_query(conn, "MATCH (n) RETURN label(n) AS label, count(*) AS rows")
```

## Export and import migration

If a future engine cannot open an old storage version, use the original engine
to export the database, then use the target engine to import into a fresh file.
Export defaults to a directory containing Parquet data and Cypher schema.

With the original engine:

```{r eval=FALSE}
old_db <- lb_database("old.lbug")
old_conn <- lb_connection(old_db)
lb_close(lb_execute(old_conn, "EXPORT DATABASE 'migration-export'"))
lb_close(old_conn)
lb_close(old_db)
```

With the upgraded package:

```{r eval=FALSE}
new_db <- lb_database("new.lbug")
new_conn <- lb_connection(new_db)
on.exit({
  lb_close(new_conn)
  lb_close(new_db)
}, add = TRUE)
lb_close(lb_execute(new_conn, "IMPORT DATABASE 'migration-export'"))
```

Use parameters for values in ordinary queries. Export and import paths are
Cypher syntax rather than value parameters, so carefully quote paths or run
these commands only with trusted, application-controlled paths.

## Incompatible-file errors

When the native engine reports a storage-version mismatch, `lb_database()`
adds an actionable message pointing to this vignette. Do not edit the database
file or retry it with concurrent writers. Restore the backup and use the
export/import workflow with an engine version that can still open the source.
